Why aren't significance tests commonly used for linear regression diagnostics?

### 2022-12-08 paper_comparison/paper_comparison/paper_comparison.tex
update text 2.3.3

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro--Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null
hypothesis for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are sensitive to
other types of departures. It is likely the null hypothesis is correctly
rejected but for the wrong reason, which is known as the ``Type III
error''. Additionally, outliers will often incorrectly trigger the
rejection of the null hypothesis despite when majority of the residuals
are well-behaved \citep{cook_applied_1999}. Furthermore, with a
sufficiently large sample size, residual-based tests may reject the null
hypothesis due to a slight departure that is of little practical
significance. These can be largely avoided in diagnostic plots as
experienced analysts can evaluate the acceptability of assumptions
flexibly, even in the presence of outliers and slight departures.

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots
\citep{gelman_exploratory_2004}. Later, \citet{buja_statistical_2009}
proposed the lineup protocol as a visual test inspired by the ``police
lineup'' or ``identity parade'' which is the act of asking the
eyewitness to identify criminal suspect from a group of irrelevant
people. The protocol consists of \(m\) randomly placed plots, where one
plot is the actual data plot, and the remaining \(m - 1\) plots have the
identical graphical production as the data plot except the data has been
replaced with data consistent with the null hypothesis. Then, an
observer who have not seen the actual data plot will be asked to point
out the most different plot from the lineup. Under the null hypothesis,
it is expected that the actual data plot would have no distinguishable
difference with the null plots, and the probability of the observer
correctly picks the actual data plot is \(1/m\). If we reject the null
hypothesis as the observer correctly picks the actual data plot, then
the Type I error of this test is \(1/m\).

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the actual data plot at position 7 is identifiable, then it
is evidence for the rejection of the null hypothesis that the regression
model is correctly specified. In fact, the actual residual plot is
obtained from a misspecified regression model with non-linearity issue.

@@

Data used in the \(m - 1\) null plots need to be simulated. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually composited
by a collection of distributions controlled by nuisance parameters.
Since regression models can have various forms, there is no general
solution to this problem, but it sometimes can be reduced to so called
``reference distribution'' by applying one of the three methods: (i)
sampling from a conditional distribution given a minimal sufficient
statistic under \(H_0\), (ii) parametric bootstrap sampling with
nuisance parameters estimated under \(H_0\), and (iii) Bayesian
posterior predictive sampling. The conditional distribution given a
minimal sufficient statistic is the best justified reference
distribution among the three \citep{buja_statistical_2009}. Essentially,
null residuals can be simulated by regressing \(N\) i.i.d standard
normal random draws on the regressors, then rescaling it by the ratio of
residual sum of square in two regressions.

@@

Further, a visual test can involve \(K\) independent observers. Let
\(D_i = \{0,1\}\) be a binomial random variable denoting whether subject
\(i\) correctly detecting the actual data plot, and
\(X = \sum_{i=1}^{K}X_i\) be the number of observers correctly picking
the actual data plot. Then, by imposing a relatively strong assumption
on the visual test that all \(K\) evaluations are fully independent,
under the null hypothesis, \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore,
the \(p\)-value of a lineup of size \(m\) evaluated by \(K\) observer is
given as \(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the
cumulative distribution function, \(f(.)\) is the probability mass
function and \(x\) is the realization of number of observers correctly
picking the actual data plot \citep{majumder_validation_2013}.

As pointed out by \citet{vanderplas2021statistical}, the binomial model
doesn't take into account the possible dependencies in the visual test
due to repeated evaluations of the same lineup. And it is inapplicable
to visual test where subjects are asked to select one or more ``most
different'' plots from the lineup. They summarized three common
scenarios in visual inference: (1) \(K\) different lineups are shown to
\(K\) subjects, (2) \(K\) lineups with different null plots but the same
actual data plot are shown to \(K\) subjects, and (3) the same lineup is
shown to \(K\) subjects. Out of these three scenarios, Scenario 3 is the
most common in previous studies as it puts the least constraints on the
experimental design. For Scenario 3, \citet{vanderplas2021statistical}
modelled the probability of a plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). And defined \(c_i\) to be the number
of times plot \(i\) being selected in \(K\) evaluations. In case subject
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots subject \(j\)
selected for \(j=1,...K\). This ensured \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the actual data plot \(i\), the
marginal model can be simplified to a beta-binomial model and thus the
visual p-value is given as

\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},
\end{equation}

@@

Note that Equation \ref{eq:pvalue-beta-binomial} only works with integer
\(c_i\). For non-integer \(c_i\), linear approximation will be applied
to calculate the p-value

\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \notin \mathbb{Z}.
\end{equation}

Besides, the parameter \(\alpha\) used in Equation
\ref{eq:pvalue-beta-binomial} is usually unknown and hence needs to be
estimated from the survey data. For low values of \(\alpha\), only a few
plots are attractive to the observers and tend to be selected. For
higher values of \(\alpha\), the distribution of the probability of each
plot being selected is more evenly. \citet{vanderplas2021statistical}
defined that a plot is \(c\)-interesting if \(c\) or more participants
selected the plot as the most different. Given the definition, The
expected number of plots selected at least \(c\) times, \(E[Z_c]\), is
calculated as

@@

\citet{vanderplas2021statistical} suggested that \(\alpha\) can be
estimated using MLE and Equation \ref{eq:c-interesting-expectation}. But
for precise estimate of \(\alpha\), additional responses to Rorschach
lineups, which is a type of lineup consists of only null plots, are
required.

@@

In particular, given 95\% confidence level, both tests will reject the
null hypothesis in only around 58\% of the cases. And in 41\% of the
cases, the visual test does not reject the null hypothesis but the
conventional test does. Since fail to reject the null hypothesis in a
visual test means that there is no obvious visual discoveries found in
the residual plot, analysts and the general public as the consumers of
the output may not fully believe in the rejection of the null hypothesis
even though the alternative hypothesis is true. Even if the rejection is
accepted, the model violation may not be considered as impactful due to
the fact that the departures are not clearly visible.

This makes it sounds like people are arrogantly ignoring the accuracy of
conventional tests and trust too much to their visual perception, but it
is not. In fact, the sensitivity of the conventional test could
sometimes distract and discourage analysts from finding a simple but
good linear approximation to the data. The rejection of the null
hypothesis because of acceptable and negligible residual departure is
not practically meaningful and useful. This also partly explains why
abundance of literature suggesting the use of data plot in regression
diagnostics.

@@

In terms of the heteroskedasticity model, the power of Breusch-Pagan
test is also almost always greater than the power of visual test. For
\(0 \leq log_e(E) \leq 2\), where the visual feature is nearly
unobservable from the example residual plots and the power curve of the
visual test remains at a low level, the Breusch-Pagan test still have a
decent amount of chance of rejecting the null hypothesis.

@@

The power of all inappropriate tests except for F-test shows improvement
as the effect increases but at a lower rate than the visual test in both
scenarios. This coincides the point made by \citet{cook1982residuals}
that residual-based tests for a specific type of model defect may be
sensitive to other types of model defects. The power curve of F-test
remains at around 5\% in Figure \ref{fig:heterpower} since there is no
non-linear term leave out in the heteroskedasticity model and the null
hypothesis of the F-test will be always satisfied.

>>>>

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro--Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null
hypothesis \(H_0\) for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are sensitive to
other types of departures. It is likely \(H_0\) is correctly rejected
but for the wrong reason, which is known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots
\citep{gelman_exploratory_2004}. Later, \citet{buja_statistical_2009}
proposed the lineup protocol as a visual test inspired by the ``police
lineup'' or ``identity parade'' which is the act of asking the
eyewitness to identify criminal suspect from a group of irrelevant
people. The protocol consists of \(m\) randomly placed plots, where one
plot is the actual data plot, and the remaining \(m - 1\) plots have the
identical graphical production as the data plot except the data has been
replaced with data consistent with \(H_0\). Then, an observer who have
not seen the actual data plot will be asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
actual data plot would have no distinguishable difference with the null
plots, and the probability of the observer correctly picks the actual
data plot is \(1/m\). If we reject \(H_0\) as the observer correctly
picks the actual data plot, then the Type I error of this test is
\(1/m\).

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the actual data plot at position 7 is identifiable, then it
is evidence for the rejection of \(H_0\) that the regression model is
correctly specified. In fact, the actual residual plot is obtained from
a misspecified regression model with non-linearity issue.

@@

Data used in the \(m - 1\) null plots need to be simulated. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggest, \(H_0\) is usually composited by
a collection of distributions controlled by nuisance parameters. Since
regression models can have various forms, there is no general solution
to this problem, but it sometimes can be reduced to so called
``reference distribution'' by applying one of the three methods: (i)
sampling from a conditional distribution given a minimal sufficient
statistic under \(H_0\), (ii) parametric bootstrap sampling with
nuisance parameters estimated under \(H_0\), and (iii) Bayesian
posterior predictive sampling. The conditional distribution given a
minimal sufficient statistic is the best justified reference
distribution among the three \citep{buja_statistical_2009}. Essentially,
null residuals can be simulated by regressing \(N\) i.i.d standard
normal random draws on the regressors, then rescaling it by the ratio of
residual sum of square in two regressions.

@@

Further, a visual test can involve \(K\) independent observers. Let
\(X_i = \{0,1\}\) be a binomial random variable denoting whether subject
\(i\) correctly detecting the actual data plot, and
\(X = \sum_{i=1}^{K}X_i\) be the number of observers correctly picking
the actual data plot. Then, by imposing a relatively strong assumption
on the visual test that all \(K\) evaluations are fully independent,
under \(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the
\(p\)-value of a lineup of size \(m\) evaluated by \(K\) observer is
given as \(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the
cumulative distribution function, \(f(.)\) is the probability mass
function and \(x\) is the realization of number of observers correctly
picking the actual data plot \citep{majumder_validation_2013}.

As point out by \citet{vanderplas2021statistical}, this basic binomial
model doesn't take into account the possible dependencies in the visual
test due to repeated evaluations of the same lineup. And it is
inapplicable to visual test where subjects are asked to select one or
more ``most different'' plots from the lineup.
\citet{vanderplas2021statistical} summarizes three common scenarios in
visual inference: (1) \(K\) different lineups are shown to \(K\)
subjects, (2) \(K\) lineups with different null plots but the same
actual data plot are shown to \(K\) subjects, and (3) the same lineup is
shown to \(K\) subjects. Out of these three scenarios, Scenario 3 is the
most common in previous studies as it puts the least constraints on the
experiment design. For Scenario 3, \citet{vanderplas2021statistical}
model the probability of a plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). And defined \(c_i\) to be the number
of times plot \(i\) being selected in \(K\) evaluations. In case subject
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots subject \(j\)
selected for \(j=1,...K\). This ensured \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the actual data plot \(i\), the
marginal model can be simplified to a beta-binomial model and thus the
visual p-value is given as

\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}

@@

Note that Equation \ref{eq:pvalue-beta-binomial} only works with
non-negative integer \(c_i\). For non-negative real number \(c_i\),
linear approximation will be applied to calculate the p-value

\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}

where \(P(C \geq \lceil c_i \rceil)\) is calculated using Equation
\ref{eq:pvalue-beta-binomial} and \(P(C = \lfloor c_i \rfloor)\) is
calculated by

\begin{equation}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter \(\alpha\) used in Equation
\ref{eq:pvalue-beta-binomial} is usually unknown and hence needs to be
estimated from the survey data. For low values of \(\alpha\), only a few
plots are attractive to the observers and tend to be selected. For
higher values of \(\alpha\), the distribution of the probability of each
plot being selected is more even. \citet{vanderplas2021statistical}
define that a plot is \(c\)-interesting if \(c\) or more participants
select the plot as the most different. Given the definition, The
expected number of plots selected at least \(c\) times, \(E[Z_c]\), is
calculated as

@@

\citet{vanderplas2021statistical} suggest that \(\alpha\) can be
estimated using MLE with Equation \ref{eq:c-interesting-expectation}.
But for precise estimate of \(\alpha\), additional responses to
Rorschach lineups, which is a type of lineup consists of only null
plots, are required.

@@

In particular, given 95\% confidence level, both tests will reject
\(H_0\) in only around 58\% of the cases. And in 41\% of the cases, the
visual test does not reject \(H_0\) but the conventional test does.
Since fail to reject \(H_0\) in a visual test means that there is no
obvious visual discoveries found in the residual plot, analysts and the
general public as the consumers of the output may not fully believe in
the rejection of \(H_0\) even though the alternative hypothesis is true.
Even if the rejection is accepted, the model violation may not be
considered as impactful due to the fact that the departures are not
clearly visible.

This makes it sounds like people are arrogantly ignoring the accuracy of
conventional tests and trust too much to their visual perception, but it
is not. In fact, the sensitivity of the conventional test could
sometimes distract and discourage analysts from finding a simple but
good linear approximation to the data. The rejection of \(H_0\) because
of acceptable and negligible residual departure is not practically
meaningful and useful. This also partly explains why abundance of
literature suggesting the use of data plot in regression diagnostics.

@@

In terms of the heteroskedasticity model, the power of Breusch-Pagan
test is also almost always greater than the power of visual test. For
\(0 \leq log_e(E) \leq 2\), where the visual feature is nearly
unobservable from the example residual plots and the power curve of the
visual test remains at a low level, the Breusch-Pagan test still have a
decent amount of chance of rejecting \(H_0\).

@@

The power of all inappropriate tests except for F-test shows improvement
as the effect increases but at a lower rate than the visual test in both
scenarios. This coincides the point made by \citet{cook1982residuals}
that residual-based tests for a specific type of model defect may be
sensitive to other types of model defects. The power curve of F-test
remains at around 5\% in Figure \ref{fig:heterpower} since there is no
non-linear term leave out in the heteroskedasticity model and \(H_0\) of
the F-test will be always satisfied.
